\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-02 paper.tex
Peters comments
closes #424

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Other subtly different concepts that we must be careful to distinguish
are the ``true'' ARG, % I don't see a better word for "true" here
describing the actual history of a sample of genomes,
and a ``population'' ARG which is
the true ARG of every individual in a population.
% Note: I started off talking about the "ultimate" ARG of everone
% who ever lived, but didn't find it that helpful.
True ARGs are the actual history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Although population ARGs unquestionably exist,
they can also never be entirely known because of a fundamental
lack of mutational information.
Even if mutation rate were high enough to uniquely identify every
recent branching point,
such a high rate would saturate the genome with mutations
and obscure deeper history.

% Population ARGs are here, now.
Population ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
and exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of this cohort.
UKB is just one of many such population-scale sequencing
projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}.
Agricultural datasets are on a similar scale,
and also include dense multi-generational sampling
and near-perfect pedigree
information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
Recent advances have made it possible to actually
\emph{estimate} ARGs at population scale:
ARGs have already been inferred for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
While this new population-scale reality presents many
exciting opportunities, it also poses substantial challenges
to existing methodologies.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG. It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
backwards in time. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

This combinatorial approach provides at least
one novel insight, clarifying the fundamental relationship between
ARGs and local trees.
Suppose we are given a gARG defined by a set of nodes and edges.
There is no requirement on the structure of this ARG beyond
the basic definitions: it could correspond to an ARG in which
every recombination event is exactly specified
(e.g.\ Fig.~\ref{fig-ancestry-resolution})
or one in which local trees are entirely disjoint
(i.e.\ only the sample nodes are shared between them).
If we  are given the sequence of local trees for this gARG
encoded as an oriented tree, along with the genome interval
covered by each tree, we can recover the original gARG exactly.
% We do this site-by-site for simplicity, but hopefully it's obvious
% that this isn't necessary?
More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
Then, the edges of the ``local ARG'' for this tree is given by
$E^x = \{(u, \pi_u^x, \{x\} \mid \pi_u^x \neq 0\}$. Since the ARG
edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
the positions over which node $c$ inherits from parent $p$, we can
then simply combine the ``local ARGs'' for each position $x$
% YW: The above "combining" operation isn't obviously lossless to me:
% we could potentially lose information in the different
% combinations of intervals (e.g. if we use 2 non-combined intervals
% to represent a diamond in a 1-RE-node ARG). It seems like we might
% need an additional gARG spec to define what combinations are equivalent.
% But no-one else has commented on this, so I guess we just let it lie.
to recover precisely the same set of edges as the original ARG.
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

@@

% Indeed, a
% common statement in the literature is that modern tools such as
% \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
% construct not an ARG but ``only'' a sequence of local trees
% \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}.
% While this statement is correct inasmuch as neither method attempts to infer
% the explicit recombination events that cause adjacent trees to differ, it is
% certainly not the case that they infer independent local trees. In fact, both
% methods take advantage of genome-wide mutation information to construct
% genealogies, and as a result nodes and edges persist across multiple local
% trees. Their results can therefore be thought of as an inheritance graph, and
% represented in the gARG format.

@@

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume
mutation and recombination rates of $5.49\times10^{-9}$
and $2.40463\times10^{-9}$ per site per generation
\citep{schrider2013rates,comeron2012many}
and a constant effective population size of $1720600$ \citep{li2006inferring}, all
values being taken from the \texttt{stdpopsim} library
\citep{adrion2020community,lauterbur2023expanding}.
Software versions were \kwarg\~v1.0, \argweaverD\~(2019),
\tsinfer\~v0.3.1, and \relate\~v1.1.9: full details and replicable code for generating
these figures are available on GitHub (see Data Availability).

@@

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski and Andrew Vaughan
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Anastasia Ignatieva acknowledges support from TODO
% TODO Jerome: add EPSRC and NIH HG011395-0 grants.
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
% Anthony Wilder Wohns acknowledges support from TODO

@@

%It is important to note that parsimonious ARGs
%have quite different properties from realisations of the
%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
%Firstly, branch lengths are not typically estimated,
%% why do we need the following two lines
%and node times are required in order to compute a likelihood under the
%coalescent with recombination (see section XXX).
%Secondly, even if branch lengths were estimated,
%the realisations would have a very small likelihood since
%the coalescent with recombination is often highly \emph{un}parsimonious
%in terms of recombination events.

>>>>

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Other subtly different concepts that we must be careful to distinguish
are the ``true'' ARG, % I don't see a better word for "true" here
describing the actual history of a sample of genomes,
and a ``population'' ARG which is
the true ARG of every individual in a population.
True ARGs are the actual history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Although population ARGs unquestionably exist,
they can also never be entirely known because of a fundamental
lack of mutational information.
Even if mutation rate were high enough to uniquely identify every
recent branching point,
such a high rate would saturate the genome with mutations
and obscure deeper history.

% Population ARGs are here, now.
Population ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
along with exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of the cohort.
UKB is just one of many such population-scale sequencing
projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}.
Agricultural datasets are on a similar scale,
and also include dense multi-generational sampling
and near-perfect pedigree
information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
Recent advances have made it possible to actually
\emph{estimate} ARGs at population scale:
ARGs have already been inferred for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
While this new population-scale reality presents many
exciting opportunities, it also poses substantial challenges
to existing methodologies.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG (here, and throughout, we use the general term ``ARG'' when
the details of the specific encoding are not important).
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
backwards in time. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

This combinatorial approach provides at least
one novel insight, clarifying the fundamental relationship between
ARGs and local trees.
Suppose we are given a gARG defined by a set of nodes and edges.
There is no requirement on the structure of this ARG beyond
the basic definitions: it could correspond to an ARG in which
every recombination event is exactly specified
(e.g.\ Fig.~\ref{fig-ancestry-resolution})
or one in which local trees are entirely disjoint
(i.e.\ only the sample nodes are shared between them).
If we  are given the sequence of local trees for this gARG
encoded as an oriented tree, along with the genome interval
covered by each tree, we can recover the original gARG exactly.
% We do this site-by-site for simplicity, but hopefully it's obvious
% that this isn't necessary?
More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
Then, the edges of the ``local ARG'' for this tree is given by
$E^x = \{(u, \pi_u^x, \{x\}) \mid \pi_u^x \neq 0\}$. Because the ARG
edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
the positions over which node $c$ inherits from parent $p$, we can
then simply combine the ``local ARGs'' for each position $x$
% YW: The above "combining" operation isn't obviously lossless to me:
% we could potentially lose information in the different
% combinations of intervals (e.g. if we use 2 non-combined intervals
% to represent a diamond in a 1-RE-node ARG). It seems like we might
% need an additional gARG spec to define what combinations are equivalent.
% But no-one else has commented on this, so I guess we just let it lie.
to recover precisely the same set of edges as the original ARG.
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

@@

@@

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume
mutation and recombination rates of $5.49\times10^{-9}$
and $2.40463\times10^{-9}$ per site per generation
\citep{schrider2013rates,comeron2012many}
and a constant effective population size of $1720600$ \citep{li2006inferring}, all
values being taken from the \texttt{stdpopsim} library
\citep{adrion2020community,lauterbur2023expanding}.
Software versions were \kwarg~v1.0, \argweaverD~(2019),
\tsinfer~v0.3.1, and \relate~v1.1.9.
Full details and code for generating
these figures are available on GitHub (see Data Availability).

@@

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski, Andrew Vaughan
and Peter Ralph for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Anastasia Ignatieva acknowledges support from TODO
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
% Anthony Wilder Wohns acknowledges support from TODO

@@
